---
title: "DS5220 Project 1 "
author: "Varsha Ranganathan"
date: "02/03/2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions

R has excellent visualization capabilities, especially with the `ggplot2` package.  Please read Chapter 3 of  [*R for Data Science [GW],* Garrett Grolemund, Hadley Wickham](https://r4ds.had.co.nz), and complete the exercises below after you finish each section.  Edit the markdown file which came with this html directly.  Make sure to enter your R code in the chunks following each question to demonstrate your answers.  Follow each code block with a text description of your solution.  Answers without demonstration will be given little credit.  Code with no description (if requested) will be given little credit.

## 3.2.4 Exercises

1. Run ggplot(data = mpg). What do you see?  

Displays an empty plot. This is because the geom function and the x and y 
mappings are not specified

```{r problem 3.2.4.1}
library(tidyverse)
ggplot(data = mpg)
```


2. How many rows are in mpg? How many columns? Demonstrate how you obtained your answers using R.

There are 234 rows and 11 columns

We can use the nrow(data) and the ncol(data) functions
We can also use the glimpse function to find the number of rows and columns
Below is a demonstration of all three functions

```{r problem 3.2.4.2}
nrow(mpg)
ncol(mpg)

glimpse(mpg)

```



3. What does the `drv` variable describe? Read the Help Panel in RStudio  by typing `?mpg` in the Console Panel to find out.  (You will see no output from RMarkdown here.)  Produce a description of `drv` by typing `mpg` below.


```{r problem 3.2.4.3}
?mpg
drv_description <- "the type of drive train, where f = front-wheel drive, r = rear wheel drive, 4 = 4wd"
cat("Description of 'drv' variable:\n", drv_description)
```




4. Make a scatterplot of `hwy` vs `cyl` using `geom_point`.

```{r problem 3.2.4.4}
ggplot(data=mpg) +
  geom_point(mapping = aes(x = hwy, y = cyl))
```

5. What happens if you make a scatterplot of `class` vs `drv`? Why is the plot not useful?

A scatter plot is not a useful display of these variables since both drv and 
class are categorical variables. Since categorical variables typically take 
a small number of values, there are a limited number of unique combinations 
of (x, y) values that can be displayed. 
In this data, drv takes 3 values and class takes 7 values, meaning that there 
are only 21 values that could be plotted on a scatterplot of drv vs. class. 
In this data, there 12 values of (drv, class) observed.
We can use the count function to find the number of unique values.

```{r problem 3.2.4.5}
ggplot(data=mpg) +
  geom_point(mapping = aes(x=class, y=drv))

count(mpg, class, drv)
```


## 3.3.1 Exercises

1. Fix the code in problem 3.3.1.1, and enter it below

Due to misplaced closing bracket. The color must be added as a parameter of 
mapping function and not the aesthetic function.

```{r problem 3.3.1.1}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color="blue")
```

2. Which variables in `mpg` are categorical? Which variables are continuous? (Hint: type `?mpg` in the Console Panel to read the documentation for the dataset in the Help Panel). How can you see this information when you run `mpg`?

When we run 'mpg', the info inside the angular <> brackets, on top of each column shows the type of the variable.

chr is categorical variable. There are 6 categorical variables.
manufacturer
model
trans
drv
fl
class

dbl and int are continuous variables. There are 5 continuous variables.
displ
year
cyl
cty
hwy

We can also use the glimpse function to get info about the mpg data from which the variable type can be identified


```{r problem 3.3.1.2}
mpg
```


3. Using your R code for 3.3.1.1, map a continuous variable to `color`, `size`, and `shape`. How does the aesthetic `shape` behave differently for mappings to `fl` and `displ`?

In the plots with color and size as the variable 'displ' we do not see an error.
In size - We can assume that the smaller points correspond to smaller values
similarly in color - darker color to smaller values
but we are unable to use a continuous variable as "shape", it shows an error. This is because shape has no hierarchy. We cannot assume that square is higher than circle or vice versa.

We are able to use the categorical variable 'fl' as shape. This does not show an error.
This is because we do not need hierarchy for categorical variables. The shapes can be random and does not have to mean any ranking or ordering.


```{r problem 3.3.1.3a}

# using continuous variable displ to aesthetic color
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color=displ))

# using continuous variable displ to aesthetic size
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size=displ))

# using continuous variable displ to aesthetic shape
# ggplot(data = mpg) +
#   geom_point(mapping = aes(x = displ, y = hwy, shape=mpg$displ), shape=mpg$displ )

# ggplot(data = mpg) +
#   geom_point(mapping = aes(x = displ, y = hwy, shape=displ))


```

Mapping `displ` to `color` and `size` produce meaningful results. Mapping `displ` to `shape` produces a warning. Assigning the aesthetic `shape` to `fl` produces the following.

```{r problem 3.3.1.3b}
# using categorical variable fl to aesthetic shape
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape=fl), color = "blue")
```


4. What happens if you map the same variable to multiple aesthetics? Use an example from your answers to problem 3.3.1.3.

Mapping the same variable to different aesthetics produces an output without any error.
although the information that is presents seems to be redundant.

```{r problem 3.3.1.4}
ggplot(data=mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color=hwy, size=displ))


ggplot(data=mpg) + 
  geom_point(mapping = aes(x = cty, y = cty, color=cty, size=cty))
```


5. What does the `stroke` aesthetic do? What shapes does it work with? (Hint: use`?geom_point`) Try it with `shape=21` and `stroke=displ` in your  code from 3.3.1.1.

Stroke defines the thickness of the border.It works for shapes 21-24.
They have both stroke color and a fill.

```{r problem 3.3.1.5}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, stroke=displ), shape=21, fill="green")
```


6.What happens if you map an aesthetic to something other than a variable name, like `aes(colour = displ < 5)`?  Try this by modifying your code in problem 3.3.1.1.

displ<5 acts as a logical expression and returns true or false.
geom_point colors the plot based on the true and false values. There is also a legend for the same.


```{r problem 3.3.1.6}

ggplot(data=mpg) +
  geom_point(mapping = aes(x=displ, y=cty, color = displ<5))

ggplot(data=mpg) +
  geom_point(mapping = aes(x=hwy, y=cty, color = displ<5))

```

